---
title: "Introduction to data visualization in R using ggplot2"
author: "WINSTATS REU"
date: "Dr. Silas Bergen"
output: html_document
---


##Overview

The R package `ggplot2` is the standard for data visualization in R.  Authored by Hadley Wickham, `gg` stands for "Grammar of Graphics."  In learning `ggplot2`, the following [cheat sheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf) will be invaluable (I myself refer to it often).  

Hadley's Grammar of Graphics is outlined in detail in [this article](http://vita.had.co.nz/papers/layered-grammar.pdf) (which I highly recommend in its entirety).  Here, he illustrates his principles using a small data set similar to the following:

```{r,}
df <- data.frame(A = c(2,1,4,9),
                 B  = c(4,1,15,80),
                 C = c(1,2,3,4),
                 D = c('far','far','near','near'))
df
```

Visualizing any data set using the Grammar of Graphics requires the use of *aesthetic mappings* that map **data** to **geoms** using **scales**.  **Scales** are the equivalent of EPTs we've studied earlier this semester.   Examples of scales in `ggplot2` are:

* **x** (position on the horizontal)
* **y** (position on the vertical)
* **color** 
* **shape**
* **size** 

There are others, but these are some important ones.  

Any plot created with `ggplot2` requires these ingredients.  To create a plot, one must specify which variable(s) is/are to be aesthetically mapped; what geom to map them to; and what scale to use to control the mapping.  The skeleton of any `ggplot2` command is as follows; parts in italics are to be replaced with specific data names/geoms/scales:


ggplot(data = *nameofdata*) + geom_*nameofgeom*(aes(*scale1* = *variable1*, *scale2* = *variable2*))



At a minimum, most geoms require the `x` scale.  

For example we could aesthetically map variable `D` using the `bar` geom.  Viewing the help page at `?geom_bar` indicates that the `x` scale is the minimum scale required to map data to the bar geom.  Other geoms have different requirements:

```{r,warning=FALSE,fig.height = 3, fig.width=4,fig.align='center'}
library(ggplot2)
ggplot(data = df) + geom_bar(aes(x = D))
```

Alternatively, we could map `A` and `B` to the `point` geom on a Cartesian plane.  Note in `?geom_point` that two scales are required for aesthetic mappings to point geoms:

```{r,warning=FALSE,fig.height = 3, fig.width=4,fig.align='center'}
ggplot(data = df) + geom_point(aes(x = A,y=B))
```

We can employ other scales outside of aesthetic mappings.  For example, if we want to change the aesthetic mapping of the above scatterplot by changing the **shape**, **color**, and **size** scale, we can do so with the following:
```{r,warning=FALSE,fig.height = 3, fig.width=4,fig.align='center'}
ggplot(data = df) + geom_point(aes(x = A,y=B), shape = 17,color='red',size=4)
```

Notice in the above code that the scales that are not controlling aesthetic data mappings are *outside* the `aes()` command.  

As an aside, looking at the `?shape` help file, we can find code to see all possible shapes:

```{r,warning=FALSE,fig.height = 3, fig.width=4,fig.align='center'}
df2 <- data.frame(x = 1:5 , y = 1:25, z = 1:25)
s <- ggplot(df2, aes(x = x, y = y))
s + geom_point(aes(shape = z), size = 4) + scale_shape_identity()
```


Now suppose we want to aesthetically map other variables using the **shape**, **color**, and **size** scales.  We must now put these specifications inside the `aes()` command and specify the variables we wish to map.  Consider the following code, and note the different looks, error messages and warnings that appear when attempting to apply aesthetic mappings using various scales depending on the data type. In `ggplot`-speak, "continuous" refers to quantitative data in general; while "discrete" refers to categorical data: 



```{r,warning=FALSE,fig.height = 4, fig.width=4.5,eval=FALSE}
#Mapping continuous C with size:
ggplot(data = df) + geom_point(aes(x = A,y=B, size = C))
#Mapping continuous C with size and color:
ggplot(data = df) + geom_point(aes(x = A,y=B, size = C, color = C))
#Mapping continuous C with shape (we know we shouldn't do this!):
ggplot(data = df) + geom_point(aes(x = A,y=B, shape = C))
```

```{r,warning=FALSE,fig.height = 4, fig.width=4.5,echo=FALSE}
ggplot(data = df) + geom_point(aes(x = A,y=B, size = C))
ggplot(data = df) + geom_point(aes(x = A,y=B, size = C, color = C))
```

**YOUR TURN**: See if you can re-create these plots:

```{r,warning=FALSE,fig.height = 3, fig.width=4.5,echo=FALSE}
#Mapping discrete D with shape:
ggplot(data = df) + geom_point(aes(x = A,y=B, shape = D), size = 4)
#Mapping discrete D with shape:
ggplot(data = df) + geom_point(aes(x = A,y=B, color = D), size = 4)
#Mapping discrete D with shape and color:
ggplot(data = df) + geom_point(aes(x = A,y=B, shape = D,color=D), size = 4)
#Mapping discrete D with shape and color:
ggplot(data = df) + geom_point(aes(x = A,y=B, shape = D,color=D,size=C))
```


Note some interesting concepts illustrated here:

* *Continuous* variables should be mapped using **size** or **color** scales; these are the scales/EPTs that can encode quantity.
* *Discrete* variables should be mapped with **shape** or **color** scales; these are the scales/EPTs that are best used for indicating "categories."  

###Layers

A very important aspect of the `ggplot2` package is the idea of *layers*.  Aesthetic mappings to different geoms can take place simply by specifying additional mappings with a `+` sign.  For example, suppose we want to create the above scatterplots with points *and* lines.  This requires two aesthetic mappings: one from the data to the points geom, and one from the data to the lines geom.  We can see this in what follows.  Note that because both `geom_point()` and `geom_line()` rely on the same aesthetic mapping, we could simplify the code by specifiying the appropriate mapping in the initial `ggplot()` command. The following two lines of code are equivalent:



```{r,warning=FALSE,fig.height = 3, fig.width=4.5,eval=FALSE}
ggplot(data = df) + geom_point(aes(x = A,y=B), size = 4) + geom_line(aes(x = A,y=B))
ggplot(aes(x = A, y = B), data = df) + geom_point(size = 4) + geom_line()
```

```{r,warning=FALSE,fig.height = 3, fig.width=4.5,echo=FALSE,fig.align='center'}
ggplot(data = df) + geom_point(aes(x = A,y=B), size = 4) + geom_line(aes(x = A,y=B))
```

**TASK:** Re-create the following plots.  What happens if you try to map variable `C` to `geom_line()` using the **size** scale?

```{r,warning=FALSE,fig.height = 3, fig.width=4.5,echo=FALSE}
ggplot(data = df) + geom_point(aes(x = A,y=B,shape=D,color=D), size = 5) + geom_line(aes(x = A,y=B),alpha=.6,size=2)
ggplot(data = df) +  geom_line(aes(x = A,y=B),size=2) + geom_point(aes(x = A,y=B,shape=D,color=D), alpha=.6, size = 5)
ggplot(data = df)  + geom_line(aes(x = A,y=B,color=D),size=2)+ geom_point(aes(x = A,y=B,shape=D), size = 4)
ggplot(data = df)  + geom_line(aes(x = A,y=B,linetype=D,color=D),size=2)+ geom_point(aes(x = A,y=B,shape=D), size = 4)
```

###Facets

*Faceting*, according to Wickham, is "a more general case of the techniques known as conditioning, trellising, and latticing, and produces small multiples showing different subsets of the data" (page 6).  This conditioning is done by way of the `facet_grid()` argument using formula, `~`, syntax:  

```{r,warning=FALSE,fig.height = 3, fig.width=4,eval=FALSE}
ggplot(aes(x = A,y=B),data = df)  + geom_line(size=2)+ geom_point(size = 4) + facet_grid(.~D)
ggplot(aes(x = A,y=B),data = df)  + geom_line(size=2)+ geom_point(size = 4) + facet_grid(D~.)
```


```{r,warning=FALSE,fig.height = 3, fig.width=4.5,echo=FALSE}
ggplot(aes(x = A,y=B, color=D),data = df)  + geom_line(size=2)+ geom_point(size = 4) + facet_grid(.~D)
ggplot(aes(x = A,y=B,color=D),data = df)  + geom_line(size=2)+ geom_point(size = 4) + facet_grid(D~.)
```

The easiest way to change the labels is to change the levels of the `D` factor:

```{r,fig.height = 3, fig.width=4,fig.align='center'}
levels(df$D)
levels(df$D) <- c('Away','Home')
ggplot(aes(x = A,y=B, color=D),data = df)  + geom_line(size=2)+ geom_point(size = 4) + facet_grid(.~D)
```


Changing the labels without changing the underlying data requires use of a `labeller` function, with is not exactly intuitive.  


### Bells and whistles

Titles and axis labels are easily modified by way of the `ggtitle()`, `xlab()`, and `ylab()` commands.  Legend titles and labels are a little less intuitive, and require thought as to the **scale** that is being shown, and the variable type that is being mapped.  Changing these is by way of the `scale_*_*()` function, where the first `*` is the scale type and the second `*` is the type of the variable being mapped.  

For example, suppose we want to change the default legend name and labels of the following graph:
```{r,warning=FALSE,fig.height = 3, fig.width=4.5,echo=FALSE,fig.align='center'}
ggplot(data = df) + geom_point(aes(x = A,y=B, shape = D,color=D), size = 4)
```

There are *two* scales involved in this visualization: **shape** and **color**, and they are being used to map a *discrete* variable.  Hence to modify the legend we should use `scale_shape_discrete()` and `scale_color_discrete()`.  Note that if we only modify one at a time, two legends will appear reflecting the different modifications; this is not desirable.  The names and labels must be identical for the two legends to merge:


```{r,warning=FALSE,fig.height = 3, fig.width=4.5,eval=FALSE}
ggplot(data = df) + geom_point(aes(x = A,y=B, shape = D,color=D), size = 4) + 
    scale_shape_discrete(name='Variable D')
ggplot(data = df) + geom_point(aes(x = A,y=B, shape = D,color=D), size = 4) + 
    scale_color_discrete(labels=c('TweedleDee','TweedleDum'))
ggplot(data = df) + geom_point(aes(x = A,y=B, shape = D,color=D), size = 4) + 
    scale_color_discrete(name='Variable D',labels=c('TweedleDee','TweedleDum')) + 
    scale_shape_discrete(name='Variable D',labels=c('TweedleDee','TweedleDum')) + 
    xlab('Variable A') + ylab('Variable B') + ggtitle("Here's a sweet title")
```


```{r,warning=FALSE,fig.height = 3, fig.width=4.5,echo=FALSE}
ggplot(data = df) + geom_point(aes(x = A,y=B, shape = D,color=D), size = 4) + 
    scale_shape_discrete(name='Variable D')
ggplot(data = df) + geom_point(aes(x = A,y=B, shape = D,color=D), size = 4) + 
    scale_color_discrete(labels=c('TweedleDee','TweedleDum'))
ggplot(data = df) + geom_point(aes(x = A,y=B, shape = D,color=D), size = 4) + 
    scale_color_discrete(name='Variable D',labels=c('TweedleDee','TweedleDum')) + 
    scale_shape_discrete(name='Variable D',labels=c('TweedleDee','TweedleDum')) + 
    xlab('Variable A') + ylab('Variable B') + ggtitle("Here's a sweet title")
```



